\documentclass[../nihongo-gakushuu-kyouzai.tex]{subfiles}
\begin{document}

\setcounter{section}{2}
\section{Essential grammar} \label{sec:essential-grammar}

Now that we have an understanding of the basics, fundamentals, and foundations of Japanese grammar, this section provides specific grammar for practical situations.

\subsection{Verb stems} \label{sec:verb-stems}

\textpurple{\emph{It seems that ``stem'' and ``root'' are used interchangeably; there is no consensus on their definitions, and different sources use either term to refer to the same concept\dots}}

\textpurple{Since there is no consensus on the definitions of root/stem, I'll use ``root'' to refer to /kik/, and ``stem'' to refer to the stem form 聞き.}

Also known as the ます-stem in other texts (we refrain from that term as the stem is used in many more conjugations, not just in ます), stems can be obtained from verbs via the rules in Table~\ref{tbl:teineigo-forms}.

Verb stems are sometimes (not always!) the noun forms of the verbs, e.g.\ \ruby{休}{やす}み is the noun form of 休む.

We can suffix the stem with target particle に or directional particle へ, then follow with a motion verb (almost always 行く or 来る). This <stem>\{に/へ\}\{行く/来る/\dots\} construct means ``to go/come do <stem>''. Use に to mean going/coming for the purpose of doing <stem>; use へ to put emphasis on the literal act of going/coming.

E.g.\ 明日、映画を\textbf{見に行く}。(Tomorrow, go to see movie.)

E.g.\ 昨日、友達が\textbf{遊びへ来た}。(Yesterday, friend came towards a playing activity.; sounds weird)

E.g.\ 昨日、友達が\textbf{遊びに来た}。 (Yesterday, friend came to play.)

The expression 「楽しみにする」 (looking forward to) is formed from a similar grammar rule (though it's a special case and should be considered a set expression).

Compound verbs can be formed by appending verbs to stems. However, there aren't formulaic rules for these; memorise them as separate verbs in their own right.

E.g.\ 走り出す (break into a run)、切り\ruby{替}{か}える (switch to something else)、\ruby{付}{つ}け\ruby{加}{くわ}える (to add something by attaching it)、\ruby{言}{い}い出す (to start talking)、言いさす (to stop talking)、読み出す (read out data/retrieve)、読みさす (to leave half-read).


\subsection{Polite forms ます、です}

The Japanese covered thus far in Section~\ref{sec:basic-grammar} is fine for five-year-olds, but adults are expected to use \ruby{丁寧語}{てい|ねい|ご} (\textbf{polite} language), \ruby{尊敬語}{そん|けい|ご} (\textbf{honorific} language) and \ruby{謙譲語}{けん|じょう|ご} (\textbf{humble} language) where appropriate.

丁寧語 is used when speaking to people of higher social rank or people you're unfamiliar with. 尊敬語 and 謙譲語 are used in professional settings, and builds upon 丁寧語.

The only indicator of which language style is in use is sentence endings. It's not possible to tell whether someone is speaking in casual or polite speech until the sentence is finished.

In 丁寧語, use \textbf{「〜ます」} and its conjugations to make verbs polite, and use \textbf{「〜です」} for nouns and adjectives (note that the polite です is incompatible with the assertive だ). Detailed rules are in Table~\ref{tbl:teineigo-forms}. Note that ます and です must come at the end of a complete sentence, and never inside any embedded/relative clauses. Refer to Appendix~\ref{appendix:conjugation-rules-summary} for a full table of 丁寧語 conjugations.

Polite verb examples:
\begin{itemize}
    \item (present-positive) 明日、大学に\textbf{行きます}。 (Tomorrow, I go to university.)
    \item (present-negative) 面白くない映画は\textbf{見ません}。(I did not see uninteresting movies.)
    \item (past-positive) 先週、ボブに\textbf{会いました}よ。 (Last week, I met Bob, you know.)
    \item (past-negative) 昼ご飯を\textbf{食べませんでした}ね。(Didn't eat lunch, right?)
\end{itemize}

Polite adjective examples:
\begin{itemize}
    \item (present-positive) 子犬はとても\textbf{好きです}。(I like puppies very much/puppies like something very much.)
    \item (present-negative) その部屋はあまり\textbf{静かじゃないです}。 (The room is not very/really quiet.)
    \item (past-positive) 先週に見た映画は、とても\textbf{面白かったです}。(The movie I saw last week was very interesting.)
    \item (past-negative) 昨日、時間が\textbf{なかったんです}。 (Yesterday, there was no time.; ある $\to$ \ul{なかった} (past negative) $\to$ なかった\ul{んだ} (explanatory) $\to$ なかったん\ul{です} (polite))
\end{itemize}

\subsubsection{です is NOT the polite form of だ, they are completely separate constructs}
This is a common misconception! です and だ are two fundamentally different concepts:
\begin{itemize}
    \item だ is declarative, whereas です is polite.

    E.g.\ そうだ is the declarative form of そう. そうです is the polite form of そう (see Table~\ref{tbl:teineigo-forms}), but not the polite form of そうだ!
    \item だ can be used both at the end of both complete sentences and relative clauses, whereas です can only be used at the end of complete sentences.

    E.g.\ そうだど思います。 \textred{(In an indirect quote, it is wrong to quote です: it must be changed to だ.)}

    E.g.\ 「はい、そうです」と答える。 (The only place where です can be in an embedded clause is in direct quotes.)
\end{itemize}

\subsection{Addressing people}
Take care to use the correct level of politeness.

\subsubsection{Referring to yourself}
To refer to yourself, use one of the following (in descending order of politeness):
\begin{itemize}
    \item \ruby{私}{わたくし}: used by both males and females, formal
    \item \ruby{私}{わたし}: used by both males and females, normal polite (you should use this most of the time)
    \item 僕: used by males, semi-polite to semi-casual
    \item 俺: used only by males, very casual, very rough
    \item あたし: used by females, cutesy and girly (most girls today use \ruby{私}{わたし} instead)
    \item your own name: used by children, feminine
    \item わし: used by elderly (males)
\end{itemize}

E.g.\ 私の\ruby{名前}{な|まえ}はジャッキーです。

\subsubsection{Referring to the listener (second-person)}
Even when directly addressing other people, you rarely use variants of ``you'', as constantly hammering the listener with ``you'' that comes across as accusatory and confrontational.

Instead, we use one of the following (in descending order of politeness):
\begin{itemize}
    \item <title>: common polite (e.g.\ \ruby{社長}{しゃ|ちょう} president/director, \ruby{課長}{か|ちょう} section manager, 先生 teacher/anyone with significant expertise including doctors)
    \item <last name><title>: common polite
    \item <last name>さん: common polite, in case no suitable title exists
    \item <first name>さん: common semi-polite
    \item <last name>くん: for males, casual/endearing, of equal or lower social position
    \item <last name>ちゃん: for females, casual/endearing, of equal or lower social position
    \item <last name>: common
    \item <first name>[\{くん, ちゃん\}]: only for people you're on first-name basis with
    \item <nothing>: common. In normal Japanese conversations, the topic/subject is commonly implied to be the listener. It's completely normal not to use anything at all, since you're directly addressing the listener!
    \item \textbf{そちら}: you, polite
    \item 君: casual, assuming/very close, used by males to address females, potentially rude
    \item あなた: \textred{rude if spoken}, use only in contexts where you must refer to the audience (e.g.\ on a physical questionnaire)
    \item \textred{THOU SHALT NOT CROSS THIS LINE}
    \item あんた: rude, assuming/familiar, expresses annoyance
    \item お前/おめえ: rude, rough and coarse, used by males
    \item てめえ: very rude, sounds like you want to beat someone up (used exclusively in movies and comics)
    \item \ruby{貴様}{き|さま}: extremely rude, sounds like you want to take someone out (used exclusively in movies and comics)
\end{itemize}

\ruby[g]{貴方}{あなた} is an old-fashioned way for women to refer to their lover or husband, nowadays only used by married middle-aged women.

\subsubsection{Referring to the third person}
For people outside of your family, use one of these:
\begin{itemize}
    \item \ruby{彼}{かれ}: he
    \item \ruby{彼女}{かの|じょ}: she
    \item 彼/ボーイフレンド: boyfriend (prefer former)
    \item 彼女/ガールフレンド: girlfriend (prefer former)
\end{itemize}

When referring to family members, there are two levels of politeness: polite and casual. We only use the casual form when we are talking about our own family members to other people. When talking about the listener's family members or when talking directly to your family members, use the polite form.

The polite form comes before the casual form:
\begin{itemize}
    \item Parents: ご\ruby{両親}{りょう|しん} / \ruby{両親}{りょう|しん}
    \item Mother: お\ruby{母}{かあ}さん / \ruby{母}{はは}
    \item Father: お\ruby{父}{とう}さん / \ruby{父}{ちち}
    \item Wife: \ruby{奥}{おく}さん / \ruby{妻}{つま}
    \item Husband: ご\ruby{主人}{しゅ|じん} / \ruby{夫}{おっと} (don't drag the trailing お vowel)
    \item Older sister: お\ruby{姉}{ねえ}さん / \ruby{姉}{あね}
    \item Older brother: お\ruby{兄}{にい}さん / \ruby{兄}{あに}
    \item Younger sister: \ruby{妹}{いもうと}さん / \ruby{妹}{いもうと}
    \item Younger brother: \ruby{弟}{おとうと}さん / \ruby{弟}{おとうと}
    \item Son: \ruby{息子}{むす|こ}さん / \ruby{息子}{むす|こ}
    \item Daughter: \ruby{娘}{むすめ}さん / \ruby{娘}{むすめ}
\end{itemize}
Yes, ご主人\ruby{様}{さま} (master/husband) is an extension of ご主人 (husband).

\subsection{The question marker か particle}
These are common question words (see Table~\ref{tbl:appendix-vocab-nouns-pronouns-and-question-words} for a more complete list):
\begin{itemize}
    \item 誰: who
    \item 何: what
    \item いつ: when
    \item どこ: where
    \item なぜ/どうして/なんで: why

    なぜ is formal and forceful\\
    どうして is softer\\
    なんで is informal (\href{https://www.reddit.com/r/LearnJapanese/comments/nxxep9/difference_between_%E3%81%A9%E3%81%86%E3%81%97%E3%81%A6_%E3%81%AA%E3%82%93%E3%81%A7_and_%E3%81%AA%E3%81%9C/}{Source})
    \item どう/どうやって: how

    どう is more general (\href{https://ja.hinative.com/questions/161399}{Source})
    \item どれ/どちら/どっち: which

    どれ: three or more\\
    どちら: two\\
    どっち: two (informal; \href{https://www.tofugu.com/japanese-grammar/kore-sore-are-dore/}{Source})
\end{itemize}

\subsubsection{か in polite questions} \label{sec:ka-in-polite-questions}
The purpose of か is to clearly mark a question in polite sentences. It's not strictly necessary, since polite sentences without a trailing か can be interpreted as a question using a rising trailing intonation during speech. However, it's commonly attached.

There is no need to use a question mark when か is used; the full-stop is used instead. Because か is polite, it is incompatible with the declarative だ.

E.g.\ お母さんはどこです\textbf{か。} \ruby{母}{はは}は買い物に行きました。(Where is your mother? My mother went shopping.; 買い物 can refer to both purchased goods and the act of shopping)

E.g.\ イタリア\ruby{料理}{りょう|り}を食べに行きません\textbf{か。} すみません。ちょっと、お\ruby{腹}{なか}がいっぱいです。(Shall we eat Italian food? Sorry, my stomach is a little full.; いっぱい: full; Note that the question is phrased in the negative, see \S\ref{sec:positive-negative-questions})

\subsubsection{Positive and negative polite questions} \label{sec:positive-negative-questions}
\emph{See discussion at \href{https://ja.hinative.com/questions/16031339}{HiNative}.}

Positive polite questions are plain questions; negative questions have a nuance of suggestion/invitation.
\begin{itemize}
    \item 食べに行き\textbf{ます}か？: Are you going to eat?
    \item 食べに行き\textbf{ません}か？: Shall we go eat?
\end{itemize}

\subsubsection{か in casual questions: binary and sarcasm}
か has a slightly different purpose in casual speech. Casual questions usually either use explanatory 「の？」 or nothing at all, so か is not used here to craft questions. Instead, it is used specifically to:
\begin{itemize}
    \item question whether something is true or not

    E.g.\ こんなのを本当に食べる\textbf{か}？ (This kind of thing, will they really eat?; こんなの is こんな + の (possession), where こんな means this type of)

    \item make rhetorical questions/express sarcasm

    E.g.\ そんなのは、ある\textbf{か}よ！ (That kind of thing, do  I look like I would have something like that!?; そんなの is そんな + の (possession), where そんな means that type of)
\end{itemize}
Most actual questions use explanatory の or nothing at all apart from a rise in intonation.

E.g.\ こんなのを本当に食べる？ (Something like this, are you really going to eat?)

E.g.\ そんなのは、ある\textbf{の}？ (Do you have something like that?)

\subsubsection{か in embedded clauses: referring to embedded questions} \label{sec:ka-question-embedded-clauses}
This functions similar to direct quoting, and marks the questions in an embedded clause. The outer clause can then talk about the embedded question.

E.g.\ \textbf{昨日何を食べたか}忘れた。 (What I ate yesterday, I forgot.)

E.g.\ \textbf{彼は何を言ったか}分からない。 (What he said, I don't understand.)

E.g.\ \textbf{先生が学校に行ったか}教えない？ (Whether the teacher went to school (binary question), would you please inform me (invitation)?)

To ask ``whether or not'' (binary question), we can use either <positive>か\textbf{どうか}, or <positive>か<negative>か.

E.g.\ 先生が学校に\textbf{行ったか行かなかった}か知らない。(Whether the teacher went to school or not, I don't know.)

E.g.\ 先生が学校に\textbf{行ったか\ul{どうか}}か知らない。(Whether the teacher went to school or not, I don't know.)


\subsubsection{Modifying question words with suffixes か、も、でも}
「」か refers to a particular existence (some\textasciitilde), 「」も refers to the universal (every\textasciitilde), and 「」でも refers to a non-particular existence (any\textasciitilde) (not to be confused with でも for ``but''). Question words and their variants are showed in Table~\ref{tbl:question-word-modifications}.

% Help: \SetCell[r=2,c=2]{c,m} <content>, \cmidrule[l]{3-4}
% Help: colspec: X[ratio, horizontal alignment] columns grow to fit width=\linewidth
%                  negative ratios: shrink to fit content and may not grow to full ratio
% Help: colspec: l/c/r columns do not grow
\longtabse[0.75]  % scale factor
{Question words and their modified variants. Treat these all as normal nouns. $^\dagger$: 誰も is usually used in negative sentences to mean nobody can do the verb, and to express the positive universal everybody we typically use \ruby{皆}{みんあ}[さん]; $^\ddagger$: 何も is used exclusively in negative sentences.}  % caption
{tbl:question-word-modifications}  % label
{
    colspec={X[-1,r]X[-1,c]X[-1,r]X[-1,c]X[-1,r]X[-1,c]X[-1,r]X[-1,c]},
    rowhead=1,
    % width=\linewidth,  % useful only with X columns
}  % inner specification options
{
    \toprule
    \textbf{「」} & \textbf{Meaning} & \textbf{「」か} & \textbf{Meaning} & \textbf{「」も} & \textbf{Meaning} & \textbf{「」でも} & \textbf{Meaning} \\
    \toprule
    誰 & who & 誰か & someone & 誰も & nobody$^\dagger$ & 誰でも & anybody \\
    \ruby{何}{なに} & what & \ruby{何}{なに}か & something & \ruby{何}{なに}も & nothing$^\ddagger$ & \ruby{何}{なん}でも & anything \\
    いつ & when & いつか & sometime & いつも & always/never & いつでも & anytime \\
    どこ & where & どこか & somewhere & どこ[に]も & everywhere/nowhere & どこでも & anywhere \\
    なぜ & why & なぜか & some reason & & & & \\
    どう & how & どうか & somehow & どれも & \emph{somehow} & どうでも & anyhow \\
    どれ & which (3 or more) & どれか & one from many & どれも & all/none & どれでも & any of many \\
    どちら & which (2) & どちらか & one of two & どちらも & both/neither & どちらでも & any of two \\
    \bottomrule
}

To mean ``For some reason...'' you can say 「どういうわけか\dots」.

「」か examples:
\begin{itemize}
    \item \textbf{誰か}がおいしいクッキーを全部食べた。 (Someone ate all the delicious cookies.)
    \item \textbf{誰が}\ruby{盗}{むす}んだのか、誰か知りませんか。 (Who stole it, doesn't anyone know?; 盗む: steal)
    \item \ruby{犯人}{はん|にん}を\textbf{どこか}で見ましたか。 (Did you see the criminal somewhere?)
    \item この\ruby{中}{なか}から\textbf{どれか}を\ruby{選}{えら}ぶの。 (You are to select a certain one from inside this.; から: from, 選ぶ: choose)
\end{itemize}

「」も examples:
\begin{itemize}
    \item この\ruby{質問}{しつ|もん}の\ruby{答}{こた}えは、\textbf{誰も}知ら\textbf{ない}。(The answer to this question, nobody knows.)
    \item 友達は\textbf{いつも}\ruby{遅}{おく}れる。 (Friend is always late.)
    \item ここにあるレストランは\textbf{どれも}おいしくない。 (All restaurants that are here are not tasty.)
    \item \ruby{今週末}{こん|しゅう|まつ}は、\textbf{どこにも}行かなかった。 (This weekend, went nowhere.; どこにも means ``target is nowhere'', も is grammatically the topic particle and should come after the target particle に, so \cancel{どこもに} is incorrect. Treat this as an exception.)
\end{itemize}

「」でも examples:
\begin{itemize}
    \item この質問の答えは、\textbf{誰でも}分かる。(The answer to this question, anyone understands.)
    \item 昼ご飯は、\textbf{どこでも}いいです。 (For lunch, anywhere is good.)
    \item あの人は、本当に\textbf{何でも}食べる。 (That person really eats anything.)
\end{itemize}

\subsection{Apologising}
\emph{Read a full article \href{https://www.clozemaster.com/blog/sorry-in-japanese/}{here}.}

Use one of the following (in descending order of politeness):
\begin{itemize}
    \item すみません: formal
    \item ごめんなさい: semi-formal
    \item ごめん[ね]: causal
    \item \ruby{悪}{わる}い: very casual, only for non-serious matters

    Can be used for past offence: \ruby{悪}{わる}かった。
\end{itemize}

\subsection{Compound sentences}

\subsubsection{て form} \label{sec:te-form}

The て-form of nouns, adjectives and verbs are used to form sequences of states (nouns/adjectives) or actions. The conjugation rules are in Appendix~\ref{appendix:conjugation-rules-summary}.

て-form conjugations for です/ます/ません exist but they are part of 尊敬語 and are outside the scope of 丁寧語. In a chain of nouns/adjectives/verbs, only the last one takes the polite form です/ます/ません; everything before takes the plain form. Since て is used when connecting sentences, there is no need for です/ます/ません to have a て-form in 丁寧語.

\subsubsection{Compound sentences (chain of descriptors/actions) using て/で and て-form}
Similarly to how we can join ``My room is clean. It is quiet. I like it a lot.'' into ``My room is clean, quiet, and I like it a lot.'' in English using the ``and'' connector, we can do the same using ``て/で'' as a connector. Just as と is the noun connector (in exclusive listings), て/で is the connector for nouns, adjectives and verbs.

The syntax is [<て-form>て/で]$^*$[<plain/polite form>]. て/で functions like a connector (between two nouns/adjectives/verbs). The entity that comes before is turned into the て-form. The final one in the list does NOT use て-form (use either plain or polite), and determines the tense (present/past) of the entire chain.

E.g.\ \ruby{食堂}{しょく|どう}に\textbf{行って}、昼ご飯を\textbf{食べて}、昼寝を\textbf{する}。 (I will go to canteen, eat lunch, and take a nap.)

E.g.\ \ruby{食堂}{しょく|どう}に\textbf{行って}、昼ご飯を\textbf{食べて}、昼寝を\textbf{した}。 (I went to canteen, ate lunch, and took a nap.)

E.g.\ 時間があって、映画を見ました。 (There was time, and I watched a movie.)

\subsubsection{Causation and reasoning particles から、ので} \label{sec:causation-reasoning-particles}
To express the direct causation (because) and reasoning (therefore) relationships, use から and ので particles:
\begin{itemize}
    \item から: direct cause marker particle (also from-marker: \S\ref{sec:verb-particles}; see also Table~\ref{tbl:appendix-vocab-adverbs-grammatical}). The possible syntaxes are:
    \begin{itemize}
        \item <direct cause>[だ]から<result>

        \textred{If the cause is a non-conjugated noun or な-adjective, you must add だ to differentiate it from the from-marker usage of から.}

        E.g.\ 時間がなかった\textbf{から}パーチィーに行きませんでした。 (There was no time, so didn't go to party.)

        E.g.\ 友達\textbf{から}プレゼントが来た。 (Present came from friend.; from-marker usage)

        E.g.\ 友達\textbf{だから}プレゼントが来た。 (Present came because of friend.; cause marker usage; this sentence sounds a bit odd)

        \item{} だから<result>

        The cause can be omitted if clear from context. \textred{Here, だ is compulsory.}

        E.g.\ 時間がなかった。\textbf{だから}パーチィーに行かなかったの？ (I didn't have time. Is that why you didn't go to the party?)
        \item <direct cause>[だ]から[です]

        The result can be omitted if clear from context. から can be treated as a regular noun, so in polite speech, add the です suffix.

        E.g.\ どうしてパーチィーに行きませんでしたか？時間がなかった\textbf{から}です。 (Why didn't you go to the party? Because I didn't have time.; polite)

        E.g.\ パーチィーに行かなかったの？うん、時間がなかった\textbf{から}。(You didn't go to the party? Yeah, because I didn't have time.; casual)
    \end{itemize}
    \item ので: non-causal explanation/reason marker, carries flavour of explanatory-の. Similar usage patterns as から (almost interchangeable), but less binding than から in that \ul{ので doesn't assert that the marked reason is the \emph{direct cause} of the result}. ので is thus softer and more polite, and preferred when explaining a reason for doing something considered discourteous.

    Because で is involved, ので is a connector that must exist between two sentences (if dangling, the sentence is implied to be a dangling sentence).
    \begin{itemize}
        \item <reason>[な]ので<result>.

        \textred{If the cause is a non-conjugated noun or な-adjective, you must add な to differentiate it from the possession marker usage of の.}

        E.g.\ 時間がなかった\textbf{ので}パーチィーに行きませんでした。 (There was no time, therefore didn't go to party.)

        E.g.\ ちょっと忙し\textbf{ので}、そろそろ\ruby{失礼}{しつ|れい}します。 (I'm a little busy, therefore I'll be making my leave soon.; 失礼します literally means ``I'm doing a discourtesy'' and is used to politely mean you're make your leave or disturbing someone's time.)

        E.g.\ 私は学生\textbf{\textred{な}\{の/ん\}で}、お金がないんです。 (I am a student, therefore I have no money.)

        E.g.\ ここは静か\textbf{\textred{な}\{の/ん\}で}、とても\ruby{穏}{おだ}やかです。 (It is quiet here, therefore it is very calm here.)
    \end{itemize}

    If omitting the reason or result (which is clear from context), use the explanatory-の particle instead (\textbf{の[だ]/のです/んだ/んです} see \S\ref{sec:noun-related-particles}).

    \begin{itemize}
        \item な\{の/ん\}で<result>

        \textred{Here, な is compuslory.}

        E.g.\ \textbf{\textred{な}\{の/ん\}}で、友達に会う時間がない。(Therefore, there is no time to meet friend.)

        \item <reason>\{の/のだ/のです/んだ/んです\}

        E.g.\ どうして\ruby{遅}{おく}れた\textbf{\{のです/んです\}}か？ (Why were you late?)

        E.g.\ パーチィーに行かなかったの？うん、時間がなかった\textbf{\{の/のだ/んだ\}}。(You didn't go to the party? Yeah, because I didn't have time.; casual)

        E.g.\ どうしてパーチィーに行きませんでしたか？時間がなかった\textbf{\{のです/んです\}}。 (Why didn't you go to the party? Because I didn't have time.; polite)
    \end{itemize}

    % E.g.\ 時間がなかった。それはパーチィーに行かなかった\textbf{\{の/ん\}です}か？ (I didn't have time. Is that why you didn't go to the party?; polite)
\end{itemize}

\subsubsection{Despite marker particle のに} \label{sec:despite-marker-particle}
To express the idea of ``despite'', the のに marker is used. The schema is <despite>のに、<sentence>.

E.g.\ 毎日運動した\textbf{のに}、全然\ruby{痩}{や}せなかった。 (痩せる: to slim)

\textred{Note that non-conjucated state-of-being nouns and na-adjectives must be tagged with the な particle, similar to rules for explanatory の.}

E.g.\ 学生\textbf{\textred{な}のに}、彼女は勉強しない。

\subsubsection{General and contradiction connector particles けど(、けれど、けれども)、が} \label{sec:general-and-contradiction-connector-particles}
けど and が are used as general connectors of any two sentences, like how we construct running sentences in English using ``and''. Also, they can be used to express the idea of contradiction between the two sentences: in this usage, が is slightly more polite (stronger contradiction) than けど. Politer forms of けど are けれど and けれども (\S\hl{???}). The schema is <sentence 1>が/けど/けれど/けれども、<sentence 2>.

E.g.\ マトリックスを見た\textbf{けど}、面白かった。 (general connector; I watched ``The Matrix'' and it was interesting.)

E.g.\ デパートに行きました\textbf{が}、いい物がたくさんありました。 (general connector; I went to the department store and there was a lot of good stuff.)

E.g.\ デパートに行きました\textbf{が}、何も欲しくなかったです。 (contradictory connector; I went to the department store but there was nothing I wanted.)

\textred{Note that non-conjucated state-of-being nouns and na-adjectives must be tagged with the だ state-of-being assertion.}

E.g.\ 今日は暇\textbf{\textred{だ}けど}、明日は忙しい。

E.g.\ \textbf{\textred{だ}けど}、彼がまだ好き\textred{\textbf{な}}の。 (That may be so, but I still like him.)

\subsection{Reason vague listing connector し} \label{sec:reason-vague-listing-connector}
し is used to list reasons for a verb or state-of-being. It is vague in the same sense as the や particle (\S\ref{sec:noun-related-particles}): there is a nuance that there may be other reasons not listed. The schema is (<reason>し)*<reason>.

E.g.\ どうして彼が好きなの？優しい\textbf{し}、かっこいい\textbf{し}、面白いから。(から means because here.)

\textred{Note that non-conjucated state-of-being nouns and na-adjectives must be tagged with the だ state-of-being assertion.}

E.g.\ どうして友達じゃないんですか？先生\textbf{\textred{だ}し}、\ruby{年上}{とし|うえ}\textbf{\textred{だ}し}\dots

For a less vague, more closed listing, use the て-form instead.

E.g. どうして彼が好きなの？優しくて、かっこよくて、面白いから。

\subsection{Adjective and verb vague listing construct 〜たり[です]} \label{sec:adj-verb-vague-listing-construct}
This is the verb/adjective version of the や particle (\S\ref{sec:noun-related-particles}). For each verb/adjective in the sequence, conjugate to past tense and add 「り」. Additionally, for the final one, tag on a 「する」, which will control the tense (\{positive, negative\} \times \{present, past\}) of the entire sentence. The schema is (<past adj/v>り、)*<past adj/v>りする.

E.g.\ 映画を見た\textbf{り}、本を読んだ\textbf{り}、昼寝した\textbf{りする}。(I do things like watch movies, read books, and take naps (among other things).)

E.g.\ この大学の授業は簡単だった\textbf{り}、難しかった\textbf{りです}。(Classes in this university are sometimes easy, sometimes difficult (and sometimes other descriptors).)

The tense of the entire sentence can be changed by conjugating the trailing する.

E.g.\ 映画を見た\textbf{り}、本を読んだ\textbf{り}、昼寝した\textbf{りしない}。(I don't do things like watch movies, read books, and take naps (among other things).)

E.g.\ 映画を見た\textbf{り}、本を読んだ\textbf{り}、昼寝した\textbf{りした}。(I did things like watch movies, read books, and take naps (among other things).)

E.g.\ 映画を見た\textbf{り}、本を読んだ\textbf{り}、昼寝した\textbf{りしなかった}。(I didn't do things like watch movies, read books, and take naps (among other things).)


\subsection{Progressive form 〜ている (enduring state of action, enduring state-of-being)} \label{sec:progressive-form}

To express an enduring state of action (-ing in English), conjugate the verb to the て-form (\S\ref{sec:te-form}, then append 〜ている. This form can be used regardless of whether the verb subject is animate or inanimate. Further conjugations of the progressive form follow rules of the いる verb.

\emph{Interestingly, ある does not have a progressive form. See discussion in Section~\ref{sec:potential-form}.}

E.g.\ 友達は何を\textbf{している}の？昼ご飯を\textbf{食べている}。 (What is friend doing? Eat\ul{ing} lunch.; している is the progressive form of する.)

E.g.\ 何を\textbf{読んでいます}か。教科書を\textbf{読んでいます}。 (います is the polite form of いる.)

E.g.\ 話を\textbf{聞いています}か。ううん、\textbf{聞いていません}。 (いません is the polite form of いない.)

In casual speech (not applicable in writing!), 〜ている can be shortened to 〜てる. However, the い is not dropped in the polite progressive form 〜ています. \textred{Try not to let this casual speak become a habit too early on! Learn the correct form first.}

E.g.\ 友達は何を\textbf{してる}の？昼ご飯を\textbf{食べてる}の。 (食べてる = eating (casual); 食べる = eat (dict))

In fact, laziness can go further: 何をしているの can be shortened by dropping particles and unnecessary vowels to become 何してんの? A related lazy casual expression is 「何やってんの」, which is a contraction of 「何をやっているの」.

The same form can also be used to convey \textbf{enduring state-of-being}, after a verb is performed. \textbf{This is decided by context.} E.g.\ 結婚している can either mean someone is currently getting married, or that someone is married and is currently in the married state. \textbf{This becomes less context-based and more of a rule for the verbs 知る、分かる, and motion verbs.}

\begin{itemize}
    \item \textbf{知っている} means to be/have been in the enduring \emph{state} of knowing. \textred{知る refers to the \emph{change} of state from not knowing to knowing, it's different and \ul{rarely used in its unconjugated form}!} (\href{https://ja.hinative.com/questions/258648}{[HN1]}, \href{https://ja.hinative.com/questions/10745394}{[HN2]})

    E.g.\ \ruby{道}{みち}は\textbf{知っています}か？ はい、今日に知りました。 (Do you know the way? Yes, I found out about it today.)

    E.g.\ この歌を\textbf{知っています}か。はい、１年前から\textbf{知っています}よ。 (Do you know this song? Yes, I've known it for a year now.)

    \item \textbf{わかっている} means to be/have been in the enduring \emph{state} of understanding. わかる refers to the \emph{change} from not understanding to understanding, it's different! (\href{https://ja.hinative.com/questions/355185}{[HN]})

    E.g.\ \ruby{前方}{ぜん|ぽう}は危ない、わかりますか。はい、はい、\textbf{わかりました}。 (The area in front is dangerous, do you understand? Yes, yes, I am aware of that.)

    E.g.\ \ruby{前方}{ぜん|ぽう}は危ない、わかりますか。はい、はい、\textbf{わかっています}。 (The area in front is dangerous, do you understand? Yes, yes, I am already aware of that, you don't have to tell me.)

    \textred{わかっている, in certain contexts, can make you sound proud or dismissive. Be careful!}
    \item \textbf{<motion verb て-form>ている} means to <motion verb> and exist (<v> + いる), and refers to the state-of-being after performing the verb.

    E.g.\ \ruby[g]{鈴木}{すずき}さんはどこですか。もう、\ruby{家}{いえ}に\textbf{帰っている}。 (Suzuki-san is already at home.)

    E.g.\ \ruby[g]{美恵}{みえ}ちゃんはどこですか。もう\textbf{来ている}よ。(行って is the て-form of 行く; 来て is the て-form of 来る, ``来って'' is wrong!)
\end{itemize}


\subsection{Resultant state 〜てある} \label{sec:resultant-state-tearu}
To express that an action is completed, with the implicit nuance that the action was completed in preparation for something else, we use 〜てある.

E.g.\ \ruby{準備}{じゅん|び}どうですか？準備は、もう\textbf{してある}よ。

E.g.\ \ruby{旅行}{りょ|こう}の\ruby{計画}{けい|かく}は終わった？うん、切符を買ったし、ホテルの\ruby{予約}{よ|やく}も\textbf{してある}。

\subsection{The auxiliary verb \ruby{置}{お}く in 〜ておく、〜とく} \label{sec:auxiliary-verb-teoku}
To express that an action is completed (or will be completed), with the explicit nuance that the action was completed in preparation for something else, we use the auxiliary verb 〜ておく. \ruby{置}{お}く when used directly (meaning ``to put/place'') is written in 漢字, but when used as an auxiliary verb it is written in 仮名 alone. Further conjugations of the compound verb follow rules of the 置く verb.

E.g.\ 晩ごはんを\textbf{作っておく}。

E.g.\ 電池を\textbf{買っておきます}。

「〜ておく」 can be contracted to 「〜とく」. Further conjugations follow rules of the とく auxiliary verb (which means to do something in readiness for).

E.g.\ 晩ごはんを\textbf{作っとく}。

E.g.\ 電池を\textbf{買っときます}。

\subsection{Spatial and temporal directional auxiliary verbs 〜ていく、 〜てくる} \label{sec:auxiliary-verbs-teiku-tekiru}
To express that an action is spatially oriented towards or from some place, we use the 〜ていく and 〜てくる auxiliary verbs respectively. Think of them as ``<v> + go'' and ``<v> + come'' respectively. Further conjugations of the compound verb follow the rules of the trailing いく or くる verb.

E.g.\ \ruby{鉛筆}{えん|ぴつ}を学校へ\textbf{持っていく}？ (spatial; Are you taking pencil to school?)

E.g.\ 鉛筆を\ruby{家}{うち}に\textbf{持ってくる}？ (spatial; Are you bringing pencil back home?)

E.g.\ お父さんは、早く\textbf{帰ってきました}。 (spatial; きました is the polite past-tense of くる)

E.g.\ \ruby{駅}{えき}の\ruby{方}{ほう}へ\textbf{走っていった}。 (spatial; いった is the past tense of いく; \ruby{駅}{えき} is train station; \ruby{方}{ほう} is direction)

The same auxiliary verbs have a temporal usage. 〜ていく can be used to mean ``<v> continuously/progressively into the future''. 〜てくる can be used to mean ``<v> performed up to the present''.

E.g.\ 冬に\ruby{入}{はい}って、コートを来ている一が\textbf{増えていきます}。 (temporal; Entering winter, people wearing coat will gradually increase.; The 入 in \ruby{入}{はい}る is only read as い when used in compound verbs where the meaning of ``entering'' is faded.)

E.g.\ \ruby{一生懸命}{いっ|しょう|けん|めい}、\textbf{頑張っていく}！ (temporal; With all my might, I will try my hardest moving into the future!)

E.g.\ 色々な人\textbf{\ruby{付}{つ}き\ruby{合}{あ}ってきた}けど、いい人はまだ見つからない。 (temporal; Went out with various types of people up till the present, but a good person hasn't been found yet.; 付き合ってきた is the past-tense of 付き合ってくる.)

E.g.\ 日本語をずっと前から\textbf{勉強してきて}、\ruby{結局}{けっ|きょく}早めた。 (temporal; Studied Japanese from way back before, and eventually quit.; 勉強してきて is the て-form of 勉強してくる.)

\subsection{Potential form} \label{potential-form}
To express the ability to perform a verb, the verb is conjugated. できる is actually the potential form of する! For detailed conjugation rules, refer to Appendix~\ref{appendix:conjugation-rules-summary}. All potential form verbs are る-verbs. Further conjugations follow る-verb rules.

\emph{Interestingly, ある does not have a potential form. See more in Section~\ref{sec:potential-to-exist-arieru}.}

In casual/slang speech, 〜られる (for originally る-verbs and 来る) can be shortened to 〜れる, so 食べられる can be shortened to 食べれる. \textred{Try not to let this casual speak become a habit too early on! Learn the correct form first.}

E.g.\ 漢字は\textbf{書けます}か？ (Can you write Kanji?; 書けます is the polite potential form of 書く.)

E.g.\ 残念だが、今週末は\textbf{行けない}。 (行けない is the negative potential form of 行く.)

E.g.\ もう\textbf{信じられない}。 (信じられない is the negative potential form of 信じる.)

\textred{Importantly, \ul{potential forms do not have direct objects, since no actual action is taken, so を cannot be used}.} Whenever を is to be used, use が or the topic particles は、も、etc.\ instead.

E.g.\ 富士山が\textbf{\ruby{登}{のぼ}れた}。 (I could climb Mt.~Fuji.; \textred{富士山\textbf{を}登れた is wrong!})

E.g.\ 重い\ruby{荷物}{に|もつ}は\textbf{持てます}。 (I am able to hold heavy luggage.; 持てます is the polite potential form of 持つ: the potential form is 持てる.)

\subsubsection{The special cases 見られる、聞ける and the alternatives 見ることができる、聞くことができる (vs.\ 見える、聞こえる)} \label{sec:special-cases-mirareru-kikeru}
The transitive verbs 見る and 聞く are paired with their intransitive counterparts 見える and 聞こえる. However, the potential forms of 見る and 聞く, 見られる and 聞ける, exist. For these two case, expression of the ``ability to perform a verb'' meaning is left to the intransitive counterparts. The potential forms take on a spetial, additional layer of meaning: \textbf{to be given the opportunity} to see or hear something.

E.g.\ 今日は\ruby{晴}{は}れて、富士山が見える。 (It cleared up today and Mt.~Fuji is visible.; \ruby{晴}{は}れる: (of sky) to clear up)

E.g.\ 友達のおかげで、映画はタダで\textbf{見られた}。 (Thanks to my friend, I was able to watch the movie for free.; タダ: free-of-charge)

E.g.\ \ruby{周}{まわ}りがうるさくて、彼が言っていたことがあんまり聞こえなかった。 (The surroundings were noisy and I couldn't really hear what he was saying.; \ruby{周}{まわ}り: surroundings)

E.g.\ 久しぶりに彼の声が\textbf{聞けた}。 (For the first time in a long time, I was able to hear his voice.)


That said, it is actually more common to express ``given the opportunity'' as ``the seeing/hearing thing is doable (\textbf{〜\{こと/の\}ができる}) '', as in the following examples.

E.g.\ 友達のおかげで、映画はタダで\textbf{見る\{こと/の\}ができた}。 (Thanks to my friend, I was able to watch the movie for free.; タダ: free-of-charge)

E.g.\ 久しぶりに彼の声が\textbf{聞く\{こと/の\}ができた}。 (For the first time in a long time, I was able to hear his voice.)

\subsubsection{Potential to exist: あり\ruby{得}{え/う}る} \label{sec:potential-to-exist-arieru}
Interestingly, ある doesn't have a potential form. The round-about way of saying ``possible to exist'' is 「あることができる」, but this is almost never used. Instead, we use あり\ruby{得}{え/う}る, which is actually a compound verb made from ある and the auxiliary verb \ruby{得}{え}る (meaning to be able to), and not considered to be a potential form of ある by natives (see discussion on \href{https://www.reddit.com/r/LearnJapanese/comments/2wostw/potential\_\%E3\%81\%82\%E3\%82\%8B\_and\_\%E3\%81\%84\%E3\%82\%8B/}{Reddit}).

For 得, the unconjugated form can be read as う (slightly more formal and traditional), but all conjugated forms use え. \textorange{I'd stick to え always, the modern pronunciation (see discussion on \href{https://ja.hinative.com/question_summaries/109797}{[HN]}).}

E.g.\ そんなことは\textbf{あり得る}。

E.g.\ そんなことは\textbf{あり得ない}。

E.g.\ 彼が\ruby{寝坊}{ね|ぼう}したこともう\textbf{あり得る}ね。 (It is possible that he overslept.; \ruby{寝坊}{ね|ぼう}する: to oversleep)

E.g.\ 其れは、あり得ない話だよ。 (That's an impossible story/scenario.)

\subsection{〜にする、〜になる auxiliary verbs} \label{sec:auxiliary-verbs-nisuru-ninaru}
All forms in this section allow further conjugations based on the rules of 〜する and 〜なる.

For adjectives, both <adj>にする and <adj>になる have the standard interpretation of causing something to become the adjective, or to become the adjective itself respectively:
\begin{itemize}
    \item (revision and extension of \S\ref{sec:adverb}) \textbf{<na-adj>にする/<i-adj root>くする} means to make something become <adj>.

    E.g.\ 静かにする (be/make quiet)、きれいにする (make clean).

    E.g.\ 強くする (make stronger)、小さくする (make smaller).
    \item \textbf{<na-adj>になる/<i-adj root>くなる} means to become <adj>.

    E.g.\ \ruby{去年}{きょ|ねん}から\ruby{背}{せ}が\textbf{高くなった}ね。 (Your height has gotten taller from last year, huh?; \ruby{背}{せ}: height)

    E.g.\ 運動しているから、\textbf{強くなる}。 (Because I am exercising, I'll become stronger.)

    E.g.\ 勉強をたくさんしたから、\ruby{頭}{あたま}が\textbf{よくなった}。 (Because I studied a lot, I became smarter.)
\end{itemize}

For nouns, <nn>にする has a special meaning,, though <nn>になる preserves the standard interpretation:
\begin{itemize}
    \item (new!) \textbf{<nn>にする} means to decide on <nn> (``do with <nn> as destination''), and is commonlyu used when ordering things from a menu.

    E.g.\ 私は、ハンバーガーとサラド\textbf{にします}。 (I'll go with the hamburger and salad.; にします is the polite form of にする.)

    E.g.\ \ruby{他}{ほか}にいいものがたくさんあるけど、やっぱりこれ\textbf{にする}。 (There are many other good things, but as I thought, I'll go with this one.; \ruby{他}{ほか}に: besides/in addition to)
    \item \textbf{<nn>になる} means to become <nn>.

    E.g.\ 医者になる.
\end{itemize}

For verbs, while we recognise that adverbs can modify verbs, we do not have a way to convert verbs into a modifying adverb, unlike what we can do with adjectives. Instead, we refer to the verb using こと (event/thing) or \ruby{様}{よう} (manner), then tag on 〜にする or 〜になる. In fact, 〜ことにする、〜ことになる、〜ようにする、〜ようになる have special meanings (the last two are usually written in 仮名 alone).

\textorange{<v>ことに\{する/なる\} have an nuance of \emph{deciding} to <v>. <v>ようにする specifically means to \emph{try} to do <v>. <v>ようになる specifically indicates a change towards <v> which has/will take place.}

\begin{itemize}
    \item \textbf{<v>ことにする} means to decide to <v>. A related form, \textbf{<v>ことにしている}, means to decide to make it a rule/practice to <v> (``to continuously decide to <v>'').

    E.g.\ 海外に行く\textbf{ことにした}。(I decided that I will go abroad.)

    E.g.\ 私は６時に\ruby{起床}{き|しょう}する\textbf{ことにしている}。 (I make a practice of getting up at six.)

    \item \textbf{<v>ことになる} means to have been arranged/decided so that <v>.

    E.g.\ 海外に行く\textbf{ことになった}。 (It's been arranged/decided that I will go abroad.)
    \item \textbf{<v>ようにする} means to try to/make an effort to/make sure that <v>.

    E.g.\ 肉を毎日食べない\textbf{ようにする}。 (I will try not to eat meat everyday.)
    \item \textbf{<v>ようになる} means to come to be that/start to <v>. The implicit nuance is on a change of state from not <v> to <v>.

    E.g.\ 肉を毎日食べない\textbf{ようになった}。 (I started to not eat meat everyday.)
\end{itemize}

Furthermore, because potential verbs describe a state of possibility rather than an actual action, <v potential>ようになる is often used to describe a change (from a state of infeasibility) to a state of feasibility.

E.g.\ 日本に来て、スシが\textbf{食べられるようになった}。 (After coming to Japan, I became able to eat sushi.)

E.g.\ １年間練習したから、ピアノが\textbf{\ruby{弾}{ひ}けるようになった}。 (Because I practised for a year, I became able to play the piano.; \ruby{弾}{ひ}く: to play a stringed/keyboard instrument)

\textorange{All present-negative verbs end with 〜ない and can conjugate as an い-adjective!}

E.g.\ \ruby{地下}{ち|か}に入って、富士山が\textbf{見えなくなった}。 (After going underground, Mt.~Fuji became not visible.；見える \to{} 見えない \to{} 見えなくなる \to{} 見えなくなった)

\textblue{Actually, for this last example, it seems that \textbf{見えないようになる} is also acceptable.}

\subsection{Conditionals} \label{sec:conditions}
There are four ways to say ``if'' in Japanese.

\subsubsection{Conditional form 〜ば} \label{sec:conditional-form}
\emph{The Takoboto dictionary refers to this as the ``provisional form'', for some reason.}

This is the simplest, most general conditional, and is a conjugation of nouns, adjectives and verbs. Refer to Appendix~\ref{appendix:conjugation-rules-summary} for detailed conjugation rules. It means ``if <future \dots>ば, then <\dots>". \ul{The focus is on the conditional predicate}, i.e.\ the part before the ば, not on the statement that follows.

The conjugation rule for nouns and な-adjectives is suffixing 「〜であれば」, which is actually using the conditional form あれば of the verb ある. 「である」 is a formal expression we will see much later (\S\hl{???}).

In the examples that follow, the \ul{focus} is underlined.

E.g.\ 友達に\textbf{会えれば}、買い物に行きます。 (\ul{If I can meet my friend}, we will go shopping.; 会えれば is the conditional form of the potential verb 会える)

E.g.\ \ul{お金が\textbf{あれば}}いいね。 (\ul{If I had money}, it would be good, huh?)

E.g.\ \ul{\textbf{楽しければ}}、私も行く。 (\ul{If it's fun}, I'll go too.)

E.g.\ \ul{\textbf{楽しくなければ}}、私もう行かない。 (\ul{If it's not fun}, I'll also not go.)

E.g.\ \ul{\textbf{食べなければ}}病気になるよ。 (\ul{If you don't eat}, you will become sick.)

\subsubsection{Past conditional form 〜たら[ば]} \label{sec:past-conditional-form}
The past conditional form conveys the same idea as the basic conditional form (\S\ref{sec:conditional-form}), but with a difference in focus: \ul{the focus is on the statement that follows the conditional predicate}, i.e.\ the part after the たら[ば], not on the predicate itself. The optional [ば] is used for formality; in causal speech it is omitted.

The predicate always takes on the past tense, though this doesn't necessarily mean the predicate must have happened in the past. It can be a future, not-yet-happened event too, just like English's ``if \dots \emph{were} to happen/if I \emph{were} \dots, then \dots''.

In the examples that follow, the \ul{focus} is underlined.

E.g.\ \textbf{暇だったら\textlightgrey{[ば]}}、\ul{遊びに行くよ}。 (If I am free, \ul{I'll go to play}.)

E.g.\ \textbf{学生だったら\textlightgrey{[ば]}}、\ul{学生\ruby{\ul{割引}}{わり|びき}で買えます}。 (If you're a student, \ul{you can buy with a student discount.})

E.g.\ 友達に\textbf{会えたら\textlightgrey{[ば]}}、\ul{買い物に行きます}。 (If I can meet my friend, \ul{we will go shopping}.)

E.g.\ お金が\textbf{あったら\textlightgrey{[ば]}}\ul{いいね}。 (If I had money, \ul{it would be good, huh?})

There is also a second usage, to express surprise at an unexpected past outcome, via the schema <past predicate/prior event>たら\textlightgrey{[ば]}、<unexpected past outcome>.

E.g.\ \ruby{家}{うち}に\textbf{帰ったら\textlightgrey{[ば]}}、誰もいなかった。 (When I went home, there was (unexpectedly) no one there.)

E.g.\ アメリカに\textbf{行ったら\textlightgrey{[ば]}}、たくさん\ruby{太}{ふと}りました。 (As a result of going to America, I (unexpectedly) got really fat.; \ruby{太}{ふと}りました is the past polite form of \ruby{太}{ふと}る)

\textorange{The way to disambiguate between the two usages is to first recognise that in either case, when we see/hear たら[ば], we focus on what comes next. Then, if what comes next is in present tense, there's no surprise nuance; if what comes next is in past tense, then there is a surprise nuance.}

\subsubsection{Expected consequence conditionals with と} \label{sec:expected-consequence-conditionals}
\emph{This is similar in nuance to the CだたらE construct (see Table~\ref{tbl:appendix-vocab-adverbs-grammatical}), with the added element of \emph{expectedness}.}

AとB means that if A happens, then as a natural consequence B is \emph{expected} to happen. The schema is C[だ]とE. Note that the state-of-being だ must be made explicit for nouns and な-adjectives, for disambiguation with other usages of と!

E.g.\ 先生\textbf{だと}、きっと\ruby{年上}{とし|うえ}なんじゃないですか。 (If he's a teacher, surely he is senior, right?; きっと: surely; なん is the contraction of なの (noun explanatory ender), じゃないですか is a polite form of asking)

E.g.\ ボールを落とす\textbf{と}落ちる。 (If you drop the ball, it will fall.)

E.g.\ 電気を消す\textbf{と}\ruby{暗}{くら}くなる。 (If you turn off the lights, it will become dark.; \ruby{暗}{くら}い: dark)

E.g.\ 学校に行かない\textbf{と}友達と会えないよ。 (If you don't go to school, you can't meet your friends.)

E.g.\ たくさん食べる\textbf{と}太るよ。 (If you eat a lot, you will get fat.)

\subsubsection{Contextual conditionals with なら[ば]} \label{sec:contextual-conditionals}
Contextual conditionals express the nuance that the statement will occur \emph{given that} the contextual predicate holds. It's translated as ``If given <predicate>, then <statement>''. The schema is <context>なら[ば]、<statement>. The optional [ば] is used for formality; in casual speech it is omitted.

E.g.\ みんなが行く\textbf{なら\textlightgrey{[ば]}}私も行く。 (If given that everyone is going, then I'll go too.)

E.g.\ アリスさんが言う\textbf{なら\textlightgrey{[ば]}}問題ないよ。 (If given that Alice says so, then there's no problem.)

E.g.\ 図書館はどこですか。図書館\textbf{なら\textlightgrey{[ば]}}、あそこです。 (Where is the library? If given that you're talking about the library, then it's over there.)

\subsubsection{もし as an uncertainty embellishment} \label{sec:moshi-as-uncertainty-embellishment}
The adverb もし (and related ones like もしや、もしかし) are only a supplement to add the nuance of uncertainty on whether the predicate/condition is true (``perhaps/by any chance''). One of the four forms above must still be used.

E.g.\ \textbf{もし}\textbf{よかったら}、映画を見に行きますか。 (past conditional; If by any chance it's ok with you, go to watch movie?; 見に is the adverb of 見る)

E.g.\ \textbf{もし}時間がない\textbf{なら}、明日でもいいよ。 (contextual conditional; If by chance given that there's no time, tomorrow is fine as well.)

\subsection{Prohibition, requirement and permission} \label{sec:prohibition-requirement-and-permission}
Expressing the idea that something must not be done, must be done, or that it's ok to do/not do is common in Japanese.

\subsubsection{Prohibition (things that must not be done)} \label{sec:prohibition}
There are three adjectives that convey prohibition: だめ、いけない、ならない. Of the three, だめ is casual (ダメ is even more casual, used in friendly settings; \href{https://ja.hinative.com/questions/19206672}{[HN]}). Also, ならない is used more for general rules, laws and policies that apply for more than one person, so it can be interpreted as the sternest of the three. In terms of sternness, だめ $<$ いけない $<$ ならない.

いけない comes from \ruby{行}{い}けない which means bad/wrong/naughty, but when used in the prohibitory manner is written in 仮名 alone.

All conjugations of the three adjectives are shown in  Table~\ref{tbl:prohibition-conjugations}. \textblue{The polite forms taken by いけない and ならない can be explained by interpreting them as present-negative forms of the verbs いける (meaning to go well) and なる (to become) respectively.}

% Help: \SetCell[r=2,c=2]{c,m} <content>, \cmidrule[l]{3-4}
% Help: colspec: X[ratio, horizontal alignment] columns grow to fit width=\linewidth
%                  negative ratios: shrink to fit content and may not grow to full ratio
% Help: colspec: l/c/r columns do not grow
\longtabse[1]  % scale factor
{Prohibition conjugations.}  % caption
{tbl:prohibition-conjugations}  % label
{
    colspec={rlll},
    rowhead=0,
    % width=\linewidth,  % useful only with X columns
}  % inner specification options
{
    \toprule
    \textbf{Present prohibition} & だめ & いけない & ならない \\
    \textbf{Past prohibition} & だめだった & いけなかった & ならなかった \\
    \midrule
    \textbf{Present prohibition (polite)} & だめです & いけません & なりません \\
    \textbf{Past prohibition (polite)} & だめでした & いけませんでした & なりませんでした \\
    \bottomrule
}

The long-form schema for prohibition is <v te>は\{だめ/いけない/ならない\}.

E.g.\ ここに\textbf{入っては}いけません。 (You must not enter here.)

E.g.\ それを\textbf{食べては}だめ！ (You can't eat that!)

E.g.\ 夜、遅くまで\textbf{電話しては}ならない。 (You must not use the phone until late at night.)

E.g.\ 早く\textbf{寝ては}なりませんでした。 (I wasn't allowed to sleep early.)

E.g.\ \textbf{死んでは}ダメだよ！ (You can't die!)

E.g.\ \textbf{泳いでは}なりません。 (Do not swim.)

The contractions for expressing prohibition involve the following replacements. \textred{Note that the \{だめ/いけない/ならない\} must not be omitted for contracted prohibition.}
\begin{align*}
    \longoverbrace{\cdots て}{\text{v te}}は\{だめ/いけない/ならない\} &\mapsto \longoverbrace{\cdots}{\text{v te fragment}} ちゃ\{だめ/いけない/ならない\}\\
    \longoverbrace{\cdots で}{\text{v te}}は\{だめ/いけない/ならない\} &\mapsto \longoverbrace{\cdots}{\text{v te fragment}} じゃ\{だめ/いけない/ならない\}
\end{align*}
$$\text{Equivalently,}\qquad ては \mapsto ちゃ, \qquad では \mapsto じゃ\text{.}$$

\emph{The second form is used for verbs with endings in \{ぐ,む,ぬ,ぶ\}; their て-forms use で.}

The contracted schema for prohibition is <v te fragment>\{ちゃ/じゃ\}\{だめ/いけない/ならない\}.

E.g.\ ここに\textbf{入っちゃ}いけません。 (You must not enter here.)

E.g.\ それを\textbf{食べちゃ}だめ！ (You can't eat that!)

E.g.\ 夜、遅くまで\textbf{電話しちゃ}ならない。 (You must not use the phone until late at night.)

E.g.\ 早く\textbf{寝ちゃ}なりませんでした。 (I wasn't allowed to sleep early.)

E.g.\ \textbf{死んじゃ}ダメだよ！ (You can't die!)

E.g.\ \textbf{泳いじゃ}なりません。 (Do not swim.)

\emph{Side note: 「死んじゃった」 is actually a conjugated contraction of 「死んでしまう」, where しまう is an auxiliary verb meaning ``to do accidentally'', and でしまう is contracted to じゃう, then conjugated into the past tense as じゃった. It's got nothing to do with the 死んじゃ we see in this section, which comes from 死んでは.}

Finally, the ちゃ/じゃ contractions here are slightly feminine.

\subsubsection{Requirement (things that must be done)} \label{sec:requirement}
Requirement is expressed as the negation of prohibition. The <v te> is the one that is negated, and there are three grammatical schemae that can be used. All are correct, and are all used at least in formal writing, so learn them all.

The three long-form grammatical schemae for requirement:
\begin{enumerate}[label=\arabic*.]
    \item <v te-negative>は\{だめ/いけない/ならない\}.

    E.g.\ 毎日学校に\textbf{行かなくては}なりません。 (I must go to school everyday.)

    E.g.\ 宿題を\textbf{しなくては}いけなかった。 (I had to do homework.)
    \item <v negative>と\{だめ/いけない/ならない\}. This uses the expected conseuqence conditional (\S\ref{sec:expected-consequence-conditionals}). The literal meaning is: if you don't do something, the expected conseuqence is no good. \ul{This is the easiest form, and is used even outside of the expected consequence characterisation.}

    E.g.\ 毎日学校に\textbf{行かないと}なりません。 (I must go to school everyday.)

    E.g.\ 宿題を\textbf{しないと}いけなかった。 (I had to do homework.)

    \item <v conditional-negative>\{だめ/いけない/ならない\}. This uses the conditional form (\S\ref{sec:conditional-form}). For requirement, because the verb is negated, the conditional-negative always contains 〜なければ.

    E.g.\ 毎日学校に\textbf{行かなければ}なりません。 (I must go to school everyday.)

    E.g.\ 宿題を\textbf{しなければい}けなかった。
\end{enumerate}

In casual speech, however, people default to either the <v negative>と\{だめ/いけない/ならない\} or the contractions that follow. When used to express requirement, the expected conseuqence conditional と (\S\ref{sec:expected-consequence-conditionals}) loses the expected consequence nuance (it's used in more situations that what it was originally defined for).

The contractions for expressing requirement involve the following replacements. \textred{Note that the \{だめ/いけない/ならない\} is omitted for contracted requirement.}
\begin{alignat*}{2}
    [\text{first schema}] & \qquad & \longoverbrace{\cdots なくて}{\text{v te-negative}}は\{だめ/いけない/ならない\} &\mapsto \longoverbrace{\cdots な}{\text{v negative fragment}}くちゃ \\
    [\text{second schema}] & \qquad & \longoverbrace{\cdots ない}{\text{v negative}}と\{だめ/いけない/ならない\} &\mapsto \longoverbrace{\cdots ない}{\text{v negative}}と \\
    [\text{third schema}] & \qquad & \longoverbrace{\cdots なければ}{\text{v conditional-negative}}\{だめ/いけない/ならない\} &\mapsto \longoverbrace{\cdots な}{\text{v negative fragment}}きゃ
\end{alignat*}
$$\text{Equivalently,}\qquad なくて \mapsto なくちゃ, \qquad なければ \mapsto なきゃ\text{.}$$

For the first schema contraction, なくちゃ is the colloquial form of なくては, which is an abbreviation of なくてはいけない. For the third schema contraction, なきゃ is the colloquial form of なければ, which is the conditional-negative form of the verb ある. \textorange{In a sense, every verb has two colloquial ``requirement'' forms: $\cdots$なくちゃ and $\cdots$なきゃ, based on their negative forms.}

\textred{Note that these contractions cannot be conjugated into the past tense, i.e. ``〜なくちゃった'', ``〜なきゃった'', ``〜なくちゃない'', ``〜なきゃない'' do not exist! (\href{https://www.tanoshiijapanese.com/dictionary/conjugation_details.cfm?entry_id=28301&element_id=189756&conjugation_type_id=78}{TNSJ})} If you want to mean requirement \emph{in the past}, use any of the three long form schemae above. If you want to mean prohibition, use the one contracted schema in Section~\ref{sec:prohibition}. If you want to mean not a requirement (i.e.\ okay to not do), use the one schema in Section~\ref{sec:permission}.

Here are the three contracted grammatical schemae for requirement.
\begin{enumerate}[label=\arabic*.]
    \item <v negative fragment ending in な>くちゃ.

    E.g.\ 毎日学校に\textbf{行かなくちゃ}。 (I must go to school everyday.)

    E.g.\ 宿題を\textbf{しなくちゃ}。 (I \emph{have} to do homework.)
    \item <v negative>と.

    E.g.\ 毎日学校に\textbf{行かないと}。 (I must go to school everyday.)

    E.g.\ 宿題を\textbf{しないと}。 (I \emph{have} to do homework.)

    \item <v negative fragment ending in な>きゃ.

    E.g.\ 毎日学校に\textbf{行かなきゃ}。 (I must go to school everyday.)

    E.g.\ 宿題を\textbf{しなきゃ}。 (I \emph{have} to do homework.)
\end{enumerate}
(Notice that the homework example has been changed to present tense; the past tense has no contractions.)

Finally, the なくちゃ contraction here is slightly feminine.

\subsubsection{Permission (things that are okay to do or not to do)} \label{sec:permission}
The schema for permission is simply <v te>も\{いい/大丈夫/\ruby{構}{かま}わない\}. Other permissive enders work too but these three are the most common. \ruby{構}{かま}わない is the negative form of \ruby{構}{かま}う, meaning to mind.

For <v te>もいい, the も may be dropped in casual speech.

E.g.\ 全部食べて\textbf{[も]いい}よ。 (You can go ahead and it it all.; ``Even if you eat it all, it's good, you know.'')

E.g.\ 全部食べなくて\textbf{[も]いい}よ。 (You don't have to eat it all.)

E.g.\ もう帰って\textbf{[も]いい}? (Can I go home already?)

E.g.\ これ、ちょっと見て\textbf{[も]いい}？ (Can I take a quick look at this?)

E.g.\ 全部\ruby{飲}{の}んで\textbf{も大丈夫}だよ。 (It's okay if you drink it all.)

E.g.\ 全部\ruby{飲}{の}んで\textbf{も構わない}よ。 (You can drink it all, I don't mind.)


\end{document}
